---
title: Tutorial on Level Correlated Models with vector autoregressive of order 1 (VAR(1)) evolution
  of latent states
author: "Chiranjit Dutta"
date: "`r Sys.Date()`"
output: html_document
bibliography: LCMVAR.bib  
nocite: '@*'
---

In this document we demonstrate how to setup level correlated model with VAR evolution of latent states. The following set of codes can be used to reproduce our results in Table 2.

# Simulate data from LCM-VAR

We simulate a trivariate time series $Y_{j,it}$ ($~j=1,\dots,3,~i=1,\dots,30,~ t=1,\dots,500$) from an LCM-VAR model:
\begin{equation}\label{Eq: LCM-AR_sim}
\begin{aligned}
    &Y_{j,it} | \theta_{j,it}, \tau \sim \text{Gamma}\bigg(\tau, \frac{\tau}{\theta_{j,it}}\bigg),\\
    &\log(\theta_{j,it}) = x_{j,t} + \alpha_{j,it} + \beta S_{j,it}, \\
      & \boldsymbol{x}_t = \boldsymbol{\Phi} \boldsymbol{x}_{t-1} + \boldsymbol{w}_t,
\end{aligned}    
\end{equation}
where \begin{equation*}
\boldsymbol{\Phi}=\begin{pmatrix}
0.5 & 0.2 & 0.3 \\
0.4 & 0.3 & 0.5 \\
0.1 & 0 & 0.8 \\
\end{pmatrix}
\end{equation*}
and $\boldsymbol{w}_t$ is simulated from $N_{3}(\boldsymbol{0},\boldsymbol{W})$, where $\boldsymbol{W}= \text{diag}(1/20,1/25,1/20)$, i.e., $W_{1} = 1/20$, $W_{2} = 1/25$ and $W_{3} = 1/20$. The static predictor is simulated from $N(0,1)$ and the coefficient $\beta = 0.4$. The level correlated term $\boldsymbol{\alpha}_{it}$ is simulated from $N_3(\boldsymbol{0},\boldsymbol{\Sigma})$, where $\boldsymbol{\Sigma}$ is defined as \begin{equation}\label{Eq: Sigma}
\boldsymbol{\Sigma}=\begin{pmatrix}
\sigma^2_1 & \rho_{12} \sigma_1 \sigma_2 & \rho_{13} \sigma_1 \sigma_3 \\
\rho_{12} \sigma_1 \sigma_2 & \sigma^2_2 & \rho_{23} \sigma_2 \sigma_3\\
\rho_{13} \sigma_1 \sigma_3 & \rho_{23} \sigma_2 \sigma_3 & \sigma^2_3\\
\end{pmatrix},
\end{equation}
with $\sigma^2_{1} = 1/2$, $\sigma^2_{2}= 1/4$,$\sigma^2_{3} = 1/3$ and $\rho_{12} = 0.8$, $\rho_{13} = 0.7$ and $\rho_{23} = 0.6$. The responses are simulated from gamma distribution with $\tau=300$. 

```{r}
rm(list=ls())

# Setting seed for reproducibility:
set.seed(12345)

###############
# Library:
###############
library(INLA)
library(MASS)

#########################
# Length of series:
#########################
n <- 30
T <- 500
d <- 3

##################
# Parameters:
#################
tau <- 300
beta <- 0.4

# Covariance matrix \Sigma :
sigma11 <- 1/2
sigma22 <- 1/4
sigma33 <- 1/3
rho_12 <- 0.8
rho_13 <- 0.7
rho_23 <- 0.6

sigma12 <- rho_12*sqrt(sigma11)*sqrt(sigma22)
sigma13 <- rho_13*sqrt(sigma11)*sqrt(sigma33)
sigma23 <- rho_23*sqrt(sigma22)*sqrt(sigma33)

level.sigma.true <- matrix(c(sigma11,sigma12,sigma13
                             ,sigma12,sigma22,sigma23
                             ,sigma13,sigma23,sigma33),nrow=d)

phi_mat <- matrix(c(0.5,0.2,0.3,0.4,0.3,0.5,0.1,0,0.8),nrow=d,byrow=TRUE)   

# Construct W matrix:
rho.w1w2= 0
rho.w1w3= 0
rho.w2w3= 0

sigma2.w1 <- 1/20   
sigma2.w2 <- 1/25   
sigma2.w3 <- 1/20   
cov.w1w2 <- rho.w1w2*sqrt(sigma2.w1*sigma2.w2)
cov.w1w3 <- rho.w1w3*sqrt(sigma2.w1*sigma2.w3)
cov.w2w3 <- rho.w2w3*sqrt(sigma2.w2*sigma2.w3)

sigma_mat_w <- matrix(NA,nrow = d,ncol = d)
sigma_mat_w[lower.tri(sigma_mat_w)] <- sigma_mat_w[upper.tri(sigma_mat_w)] <- c(cov.w1w2,cov.w1w3,cov.w2w3)
diag(sigma_mat_w) <- c(sigma2.w1,sigma2.w2,sigma2.w3)

###########################
# Simulation of VAR(1)
###########################
sim.VAR <- function(phi_mat,n){
  #n <- T
  init_val <- MASS::mvrnorm(n = 1,mu = rep(0,d)
                            ,Sigma = sigma_mat_w)
  sample_list <- list()
  sample_list[[1]] <- init_val
  for(i in 2:n){
    sample_list[[i]] <- phi_mat %*% sample_list[[i-1]] + matrix(MASS::mvrnorm(n = 1,mu = rep(0,d)
                                                                              ,Sigma = sigma_mat_w))
  }
  
  simulated_var <- do.call(cbind,sample_list)
  var_ts <- t(simulated_var)
  colnames(var_ts) <- paste("y",1:d,sep="")
  
  return(var_ts)
}

fixed_sim_VAR <- sim.VAR(phi_mat,n=T)

# Fixed VAR process:
x_t_list <- list()
for(i in 1:n){
  x_t_list[[i]] <- data.frame(i.index=i,j.index=rep(1:d,each=T),t.index= rep(1:T,d)
                                  ,x=as.vector(fixed_sim_VAR))
}

x_t <- do.call(rbind,x_t_list)

# Generate /alpha_{it}
alpha_it_list <- list()
for(i in 1:n){
  alpha_it_list[[i]] <- data.frame(i.index=i,j.index=rep(1:d,each=T),t.index= rep(1:T,d)
                                   ,alpha=as.vector(mvrnorm(n = T,mu = rep(0,d)
                                                            ,Sigma = level.sigma.true)))
}

alpha_it <- do.call(rbind,alpha_it_list)

# Static Explanatory variable:
S_jit_mat <- data.frame(i.index=rep(1:n,each=T*d),j.index=rep(rep(1:d,each=T),n)
                        ,t.index=rep(1:T,n*d),S_jit = rnorm(d*n*T,sd=1))


theta_jit <- data.frame(i.index=alpha_it$i.index,j.index= alpha_it$j.index
                        ,t.index=alpha_it$t.index
                        ,theta=exp(alpha_it$alpha + x_t$x + beta*S_jit_mat$S_jit))

# Formatting responses:
response_jit <- data.frame(i.index=theta_jit$i.index,j.index=theta_jit$j.index
                           ,t.index=theta_jit$t.index,theta=theta_jit$theta,response=NA)

for(k in 1:nrow(response_jit)){
  response_jit$response[k] <- rgamma(n = 1,shape = tau,scale = response_jit$theta[k]/tau)
}

# Response:
all_response <- response_jit

new_comp1 <- response_jit[response_jit$j.index==1,]$response
new_comp2 <- response_jit[response_jit$j.index==2,]$response
new_comp3 <- response_jit[response_jit$j.index==3,]$response

data_response <- data.frame(new_comp1,new_comp2,new_comp3)
```

We set up indexes and data frame. 

```{r}
################################################
# Setting up indexes and  data frame for R-INLA:
################################################
N_1 <- d*T

# Index for random effects (There are N replicated d-variate random effects)
alpha_index <- rep(c(1:N_1),n)

# Index for each component 
ind_1 <- rep(c(1:T, rep(NA,T),rep(NA,T)),n)
ind_2 <- rep(c(rep(NA,T), 1:T,rep(NA,T)),n)
ind_3 <- rep(c(rep(NA,T), rep(NA,T),1:T),n)

# Replicates:
rep_alpha_index <- rep(1:n,each=N_1)
rep_ind1 <-  rep(1:n, each=N_1)
rep_ind2 <-  rep(1:n, each=N_1)
rep_ind3 <-  rep(1:n, each=N_1)

response_df_new <- data.frame(all_response,alpha_index
                              ,rep_alpha_index,ind_1,ind_2,ind_3,
                              rep_ind1,rep_ind2,rep_ind3)
```

# R-INLA setup with rgeneric:

We define the function $\textit{inla.rgeneric.VAR.model.diagW}$ using $\textit{rgeneric}$ which is a mechanism implemented in the $\textit{INLA}$ package which allows users to implement latent effects in $\textit{R}$. To define the new latent effect one has to define the following $\textit{R}$ functions:

- $\textit{mu()}$: The mean of the latent effects which in this case is a zero vector.
- $\textit{Q()}$: The precision of the latent effects,
- $\textit{initial()}$: The initial values of the parameter vector $\theta$ expressed in $\textit{R-INLA}$'s internal scale.
- $\textit{graph()}$: A graph indicating the non-zero entries of the precision matrix $\textit{Q()}$
- $\textit{log.norm.const()}$: It is log-normalizing constant which is the logarithm of the normalizing constant from a multivariate Gaussian distribution
- $\textit{log.prior()}$: This specifies the logarithm of the priors for the original model parameters
- $\textit{interpret.theta()}$: This enables us to obtain the list of original parameters and number of parameters.

```{r}
# Define previous variables as global to avoid warnings()
utils::globalVariables(c("p", "k", "N"))


'inla.rgeneric.VAR.model.diagW' <-
  function(cmd = c("graph", "Q", "mu", "initial", "log.norm.const",
                   "log.prior", "quit"), theta = NULL)
  {
    interpret.theta <- function(){
      n_phi <- k*k*p 
      n_prec <- k
      n_tot <- n_phi + n_prec
      phi_VAR <- sapply(theta[as.integer(1:n_phi)], function(x) { x })
      # W matrix of precisions     
      wprec <- sapply(theta[as.integer((n_phi+1):(n_phi+n_prec))], function(x) { exp(x) })
      param <- c(phi_VAR, wprec)
      # Initial k times k matrix with 1 on the diagonal - error cov W matrix
      W <- diag(1, n_prec)
      # Preparing the st. dev matrix
      st.dev <- 1 / sqrt(wprec)
      # Matrix of st. dev.
      st.dev.mat <- matrix(st.dev, ncol = 1) %*% matrix(st.dev, nrow = 1)
      # Final Variance-Covariance matrix: Diagonal if no correlation
      W <- W * st.dev.mat # Element wise multiplication
      # Inverting the matrix
      PREC <- solve(W)
      return(list(param = param, VACOV = W, PREC = PREC, phi_vec=c(phi_VAR)
                  ,n_phi=n_phi,n_prec=n_prec))
      
    }
    
    # Precision matrix
    Q <- function(theta){
      # Parameters:
      param <- interpret.theta()
      # Define phi_matrix and sigma_w_inverse
      phi_mat <- matrix(param$phi_vec,nrow=k)
      sigma_w_inv <- param$PREC
      # Define A,B and C matrices:
      A <- t(phi_mat)%*%sigma_w_inv%*%phi_mat
      B <- -t(phi_mat)%*%sigma_w_inv
      C <- sigma_w_inv
      # Construct Mid-block:
      zero_mat <- matrix(0,nrow=k,ncol=k*N)
      # Define the matrix block:
      mat <- cbind(t(B),A+C,B)
      # Initializing column id and matrix list:
      col_id <- list()
      mat_list <- list()
      col_id[[1]] <- 1:(3*k)
      mat_list[[1]] <- zero_mat
      mat_list[[1]][,col_id[[1]]] <- mat
      for(id in 2:(N-2)){
        start_id <- col_id[[id-1]][1]+k
        end_id <-  start_id+(3*k-1)
        col_id[[id]] <- start_id:end_id
        mat_list[[id]] <- zero_mat
        mat_list[[id]][,col_id[[id]]] <- mat
      }
      mid_block <- do.call(rbind,mat_list)
      tau_val <- 0.1
      diffuse_prec <- tau_val*diag(1,k)
      # Construct first and last row blocks and then join with mid block:
      first_row_block <- matrix(0,nrow=k,ncol=k*N)
      mat_first <- cbind(A+diffuse_prec,B)
      first_row_block[1:k,1:(2*k)] <- mat_first
      last_row_block <- matrix(0,nrow=k,ncol=k*N)
      mat_last <- cbind(t(B),C)
      last_row_block[1:k,(k*N - 2*k+1):(k*N)] <- mat_last
      toep_block_mat <- rbind(first_row_block,mid_block,last_row_block)
      # Changing to a sparse Matrix:
      library(Matrix)
      prec_mat <- Matrix(toep_block_mat,sparse = TRUE)
      return(prec_mat)
    }
    
    # Graph function: Essentially Q matrix
    graph = function() {
      return (inla.as.sparse(Q()))
    }
    
    # Mean of model
    mu <- function() {
      return(numeric(0))
    }
    
    # Log normal constant:
    log.norm.const <- function() {
      Q <- Q()
      log_det_val <- Matrix::determinant(Q, logarithm = TRUE)$modulus
      val <- (-k*N/2)*log(2*pi) + 0.5 * log_det_val
      return (val)
    }

    log.prior <- function() {
      # return the log-prior for the hyperparameters.
      param <- interpret.theta()
      pars <- param$param
      k <- k
      total_par <- param$n_phi + param$n_prec
      # Normal prior for phi's:
      theta_phi <- theta[1:param$n_phi]
      phi_prior <- sum(sapply(theta_phi,function(x) dnorm(x,mean=0,sd=1,log=TRUE)))
      theta_prec <- theta[(param$n_phi + 1):(param$n_phi+param$n_prec)]
      prec_prior <- sum(sapply(theta_prec,function(x) dgamma(x,shape = 0.1,scale = 10,log=TRUE)))
      prec_jacob <- sum(theta_prec) # This is for precision terms
      prior_val <- phi_prior + prec_prior + prec_jacob
      return (prior_val)
    }
    
    initial = function() {
      phi_init <- rep(0.1,k^2)
      prec_init <- rep(1,k)
      #rho_init <- rep(0.1, k*(k-1)/2)
      #init <- c(phi_init,prec_init,rho_init)
      init <- c(phi_init,prec_init)
      return (init)
    }
    
    # FIX for rgeneric to work on R >= 4
    # Provided by E. T. Krainski
    if (as.integer(R.version$major) > 3) {
      if (!length(theta))
        theta <- initial()
    } else {
      if (is.null(theta)) {
        theta <- initial()
      }
    }
    val <- do.call(match.arg(cmd), args = list())
    return (val)
  }


# This function allows us to obtain the list of original parameters:
interpret.theta <- function(theta){
  n_phi <- k*k*p 
  n_prec <- k
  n_tot <- n_phi + n_prec
  phi_VAR <- sapply(theta[as.integer(1:n_phi)], function(x) { x })
  #W matrix  precisions     
  wprec <- sapply(theta[as.integer((n_phi+1):(n_phi+n_prec))], function(x) { exp(x) })
  param <- c(phi_VAR, wprec)
  # initial k times k matrix with 1 on the diagonal - error cov W matrix
  W <- diag(1, n_prec)
  #Preparing the st. dev matrix
  st.dev <- 1 / sqrt(wprec)
  # Matrix of st. dev.
  st.dev.mat <- matrix(st.dev, ncol = 1) %*% matrix(st.dev, nrow = 1)
  # Final Variance-Covariance matrix: Diagonal if no correlation
  W <- W * st.dev.mat # Element wise multiplication
  # Inverting the matrix
  PREC <- solve(W)
  return(list(param = param, VACOV = W, PREC = PREC, phi_vec=c(phi_VAR)
              ,n_phi=n_phi,n_prec=n_prec))
  
}
```

# Results:

```{r}
###############################
# Setting up data for INLA :
###############################

subset_data <- response_df_new[response_df_new$i.index==1 & response_df_new$j.index==1,]
p <<- 1
k <<- d
N <<- nrow(subset_data)

##########################
# Setting index for VAR:
##########################
comp1_id <- seq(1,3*N,by=d)
comp2_id <- seq(2,3*N,by=d)
comp3_id <- seq(3,3*N,by=d)
var_id <- c(comp1_id,comp2_id,comp3_id)
response_df_new$var_id <- rep(var_id,n)

# Setting rgeneric model:
model <- inla.rgeneric.define(inla.rgeneric.VAR.model.diagW,p = p, k = k,N=N)

formula.inla= response ~ -1+f(alpha_index, model="iid3d", n=N_1
                              ,replicate = rep_alpha_index)+ S_jit_mat$S_jit +f(var_id,model=model)

result= inla(formula.inla,family="gamma",
             data =response_df_new,
             control.compute=list(dic=FALSE),
             control.family = list(hyper = list(prec = list(prior="loggamma"
                                                            ,param=c(300,1)))),
             verbose = TRUE,control.inla=list(control.vb=list(enable=FALSE)))

##################
# Results:
##################  
all_parameters_hyper <- result$summary.hyperpar
all_parameters_fixed <- result$summary.fixed

actual <- c(beta,tau,1/sigma11,1/sigma22,1/sigma33
            ,rho_12,rho_13,rho_23
            ,as.vector(phi_mat),1/diag(sigma_mat_w))

estimated_mean <- c(all_parameters_fixed$mean,
                    all_parameters_hyper$mean[1:7],
                    interpret.theta(all_parameters_hyper$mean[8:19])$param)

ci_low <- c(all_parameters_fixed$`0.025quant`,
            all_parameters_hyper$`0.025quant`[1:7],
            interpret.theta(all_parameters_hyper$`0.025quant`[8:19])$param)

ci_up <- c(all_parameters_fixed$`0.975quant`,
           all_parameters_hyper$`0.975quant`[1:7],
           interpret.theta(all_parameters_hyper$`0.975quant`[8:19])$param)

# Standard deviation of transformed variable: Var(g(x)) = (g'(/mu_x))^2 * /sigma^2_X
transformed_sd <- sqrt(exp(2*all_parameters_hyper$mean[17:19])*(all_parameters_hyper$sd[17:19])^2)
estimated_std <- c(all_parameters_fixed$sd,
                   all_parameters_hyper$sd[1:7],
                   all_parameters_hyper$sd[8:16],
                   transformed_sd)

tab <- data.frame(actual,estimated_mean,estimated_std,ci_low,ci_up)
rownames(tab) <- c("beta","tau","prec.v1","prec.v2","prec.v3"
                   ,"rho.v1.v2","rho.v1.v3","rho.v2.v3"
                   ,"phi_11","phi_21","phi_31"
                   ,"phi_12","phi_22","phi_32"
                   ,"phi_13","phi_23","phi_33"
                   ,"prec.w1","prec.w2","prec.w3")

colnames(tab) <- c("True value","Posterior Mean", "Posterior Std. Dev."
                   ,"95% CI_low","95% CI_up")

tab
```

# References


